### 2025-09-09 YEAR2/AER210_LectureNotes/main.tex
sept 9

\subsection{Formal Definition of Double Integrals}
There is two definitions of double integrals in this course, due to the discrepancy between Stewart's book and the lectures.

\begin{shaded}
    \subsubsection*{\textbf{Review.  } Formal Definition of Definite Integral (Single Variable)}

@@

\chapter{Fluid Mechanics}
\end{document}

>>>>

\subsubsection{Formal Definition of Double Integrals}
There is two definitions of double integrals in this course, due to the discrepancy between Stewart's book and the lectures.

\begin{shaded}
    \textbf{Review.  Formal Definition of Definite Integral (Single Variable)}

@@

\begin{shaded}
The following is the analogue of the squeeze theorem for double integrals:
\begin{definition}[Squeeze Theorem for Double Integrals]
    For the first defintion Consider region $R$ subdivided into $N$ subregions $R_1, R_2, \ldots, R_N$, such that all subregions $\bigcup_{i=1}^N R_i \subset R$ (They are all inside). For both cases, we require that $R_i \cap R_j = \emptyset$ for $i \neq j$, and then some of the area would be omited and the following would be garanteed:
    $$
        \sum_{i=1}^N \Delta A \le \text{Area}(R), \quad \sum_{i=1}^N m_i \Delta A_i \le \iint_R f(x,y) \, dA 
    $$ 
    where $m_i$ and $M_i$ are the minimum and maximum values of $f(x,y)$ in the $i$-th subregion. Similarly, if $\bigcup_{i=1}^N R_i \supset R$ (They all cover $R$), and that we garantee that $R_i \cap R \neq \emptyset$ for all $i$. Then, some of the area would be double counted and the following would be garanteed:
    $$
        \sum_{i=1}^N \Delta A \ge \text{Area}(R), \quad \sum_{i=1}^N M_i \Delta A_i \ge \iint_R f(x,y) \, dA
    $$ 
    For the second definition, the same logic applies, but we consider subrectangles that creates grid that is either inside or covering $R$.
\end{definition}

\end{shaded}
\begin{example}
    Estimate the volume that lies above the square $R = [0,2] \times [0,2]$ and below the surface $z = f(x,y) = 16 - x^2 - 2y^2$ by dividing the $R$ into four subrectangles of equal area and using the value of the function at the upper right corner of each subrectangle to form a Riemann sum. Choose the upper right corner of each subrectangle as the sample point.

    We divide the square $R$ into four subrectangles, each with an area of 1. We obtain the sum:
    $$
        V \approx \sum_{i=1}^2 \sum_{j=1}^2 f(x_i^*,y_j^*) \Delta A = \sum_{i=1}^2 \sum_{j=1}^2 f(i,j) \cdot 1
    $$
    where the sample points are $(1,1)$, $(1,2)$, $(2,1)$, and $(2,2)$. Evaluating the function at these points gives:
    $$
        V \approx f(1,1) + f(1,2) + f(2,1) + f(2,2) = 34
    $$
    Therefore, the estimated volume is approximately 34.
\end{example}

\subsection{Double Integrals in Non-Rectangular Regions}

\begin{theorem}[Change of Variable to Polar Coordinates]
    Consider the double integral of a function $f(x,y)$ over a region $R$ in the $xy$-plane. If we change the variables from Cartesian coordinates $(x,y)$ to polar coordinates $(r,\theta)$ using the transformations:
    $$
        x = r \cos{\theta}, \quad y = r \sin{\theta} \implies r = \sqrt{x^2 + y^2}
    $$
    then the double integral can be expressed in polar coordinates as follows:
    \begin{equation}
        \iint_R f(x,y) \, dA = \iint_{R'} f(r \cos{\theta}, r \sin{\theta}) \, r \, dr \, d\theta
    \end{equation}
    where $R'$ is the corresponding region in the $r\theta$-plane, and the term $r$ arises from the Jacobian determinant of the transformation from Cartesian to polar coordinates.
\end{theorem}
\begin{proof}
   \textbf{Note. } \textit{This change of variable can be derived using the Jacobian determinant of the transformation from Cartesian to polar coordinates, which will be covered in Section \ref{sec:change_of_variables}, which can fully prove this theorem in the case where $g < 0$ for some input.}

   \textbf{Geometric Sketch: } Assume $f(r \cos{\theta}, r \sin{\theta}) = g(r, \theta) \ge 0$. Consider a small rectangle $\Delta A_i$ in the $xy$-plane with dimensions $\Delta x_i$ and $\Delta y_i$. When we transform this rectangle into polar coordinates, it becomes a small sector of a circle with radius $r_i$ and angle $\Delta \theta_i$. The area of this sector is given by:
    $$
          \Delta A_i = r_i \Delta r_i \Delta \theta_i  \left( 1 + \frac{\Delta r_i}{2r_i} \right)
    $$
    this is derived from the geormetric formula of the area of a sector of a circle. Then, as $\Delta r_i \to 0$, the term $\frac{\Delta r_i}{2r_i} \to 0$, and we have:
    $$
        \Delta A_i \approx r_i \Delta r_i \Delta \theta_i
    $$
    Therefore, the double integral in polar coordinates can be approximated as:
    $$
        \iint_R f(x,y) \, dA \approx \sum_{i=1}^n f(r_i \cos{\theta_i}, r_i \sin{\theta_i}) r_i \Delta r_i \Delta \theta_i
    $$
    Taking the limit as the maximum diameter of the subrectangles approaches zero, we obtain the exact double integral in polar coordinates.
\end{proof}

\begin{definition}[Reigion Defined by Varying $r$ with $\theta$]
    Consider a region $R$ in the $xy$-plane that can be described in polar coordinates by the inequalities:
    $$
        \alpha \le \theta \le \beta, \quad g_1(\theta) \le r \le g_2(\theta)
    $$
    where $g_1(\theta)$ and $g_2(\theta)$ are continuous functions on the interval $[\alpha, \beta]$. Then, to evaluate the double integral over this region for a continuous function $f(x,y)$, we set up the integral as follows:
    \begin{equation}
        \iint_R f(x,y) \, dA = \int_\alpha^\beta \int_{g_1(\theta)}^{g_2(\theta)} f(r \cos{\theta}, r \sin{\theta}) \, r \, dr \, d\theta
    \end{equation}
\end{definition}

\begin{definition}[Reigion Defined by Varying $\theta$ with $r$]
    Similarly, consider a region $R$ in the $xy$-plane that can be described in polar coordinates by the inequalities:
    $$
        a \le r \le b, \quad h_1(r) \le \theta \le h_2(r)
    $$
    where $h_1(r)$ and $h_2(r)$ are continuous functions on the interval $[a, b]$. Then, to evaluate the double integral over this region for a continuous function $f(x,y)$, we set up the integral as follows:
    \begin{equation}
        \iint_R f(x,y) \, dA = \int_a^b \int_{h_1(r)}^{h_2(r)} f(r \cos{\theta}, r \sin{\theta}) \, r \, d\theta \, dr
    \end{equation}
\end{definition}

\begin{shaded}
    \textbf{When to Use Polar Coordinates}

    Polar coordinates are particularly useful for regions with circular or radial symmetry, as they simplify integration by transforming variables into a more natural form. They are also advantageous for integrands that are difficult in Cartesian coordinates, especially those involving terms like $x^2 + y^2$.
\end{shaded}

\begin{example}
    Evaluate $\iint_R (3x + 4y^2) \, dA$, where $R$ is the region in the upper half-plane bounded by the circles $x^2 + y^2 = 4$ and $x^2 + y^2 = 1$ (Donut region).
    We can describe the region $R$ in polar coordinates as $1 \le r \le 2$ and $0 \le \theta \le \pi$. Thus, we can set up the double integral as follows:
    \begin{align*}
        \iint_R (3x + 4y^2) \, dA &= \int_0^\pi \int_1^2 (3(r \cos{\theta}) + 4(r \sin{\theta})^2) \, r \, dr \, d\theta \\
        &= \int_0^\pi \int_1^2 (3r^2 \cos{\theta} + 4r^3 \sin^2{\theta}) \, dr \, d\theta \\
        &= \int_0^\pi (r^3\cos \theta + r^4 \sin^2 \theta) \Big|_{r=1}^{r=2} \, d\theta \\
        &= \int_0^\pi (7\cos \theta + 15 \sin^2 \theta) \, d\theta = \frac{15}{2} \pi \quad \left(\text{Using } \sin^2 \theta = \frac{1 - \cos 2\theta}{2}\right)
    \end{align*}
\end{example}

\begin{example}
    Find the volume of the solid bounded by the $z=0$ plane and the paraboloid $z = 1 - x^2 - y^2$. We first consider the projection of the paraboloid onto the $xy$-plane, which is the circle $1 - x^2 - y^2 = 0$. We can describe the region $R$ in polar coordinates as $0 \le r \le 1$ and $0 \le \theta \le 2\pi$. Thus, we can set up the double integral as follows:
    \begin{align*}
        V &= \iint_R (1 - x^2 - y^2) \, dA \\
        &= \int_0^{2\pi} \int_0^1 (1 - r^2) \, r \, dr \, d\theta \\
        &= \int_0^{2\pi} \left( \int_0^1 (r - r^3) \, dr \right) d\theta \\
        &= \int_0^{2\pi} \left( \frac{1}{2} - \frac{1}{4} \right) d\theta  = \frac{\pi}{2}
    \end{align*}
\end{example}

\subsection{Applications of Double Integrals}
\subsection{Surface Area}

\subsection{Triple Integrals in Rectangular Coordinates}
\subsection{Triple Integrals in Cylindrical Coordinates}
\subsection{Triple Integrals in Spherical Coordinates}
\subsection{Change of Variables in Multiple Integrals} \label{sec:change_of_variables}
\chapter{Fluid Mechanics}
\end{document}

### 2025-09-09 YEAR2/CHE260_LectureNotes/main.tex
sept 9

The first law of thermodynamics is simple and easy to understand:
\begin{definition}[First Law of Thermodynamics]
    The first law of thermodynamics states that energy cannot be created or destroyed, only transformed from one form to another. In the context of a heat engine, this means that the work output ($W$) is equal to the heat input ($Q_H$) minus the heat rejected ($Q_C$):
    \begin{equation}
        W = Q_H - Q_C
    \end{equation}
\end{definition}
The second law formalizes the concept of the directionality of thermodynamic processes, stating that heat cannot spontaneously flow from a colder body to a hotter body. This law introduces the idea of entropy, a measure of the disorder or randomness in a system, which tends to increase over time in isolated systems.

@@

\subsection{Modeling Ideal Gas}
\begin{definition}[Ideal Gas]
    We consider ideal gas conditions as those where the gas particles do not interact with each other except during elastic collisions, and the volume of the gas particles themselves is negligible compared to the volume of the container. In this course, we consider 1 atm as the threshold for ideal gas behavior. Of course, ideal gas obey the ideal gas law:
    \begin{equation}
        PV = nRT
    \end{equation}
\end{definition}

@@

\begin{definition}[Equilibrium]
    An \textbf{ISOLATED} system is in equilibrium if its properties are uniform throughout the system and do not change with time (constant). \\
    (Example) A gas in a closed container that has been left undisturbed for a long time.
\end{definition}
\begin{example}
    Consider a gas in a cylinder with a movable piston. If the piston is held in place, the gas is in a steady state, as its properties (pressure, volume, temperature) do not change with time. However, if the piston is allowed to move, the gas will expand or compress, and its properties will change with time, indicating that it is not in a steady state. 
\end{example}
\begin{definition}[Process]
    A process is a transformation from one state to another. A process can be described by the path taken by the system in the property space. For example, a gas can undergo a process of compression or expansion, which changes its pressure, volume, and temperature.
\end{definition}
\chapter{Heat Transfer}

>>>>

The first law of thermodynamics is simple and easy to understand:
\begin{definition}[First Law of Thermodynamics] \label{def:firstlaw}
    The first law of thermodynamics states that energy cannot be created or destroyed, only transformed from one form to another. In the context of a heat engine, this means that the work output ($W$) is equal to the heat input ($Q_H$) minus the heat rejected ($Q_C$):
    \begin{equation}
        W = Q_H - Q_C
    \end{equation}
\end{definition}
The second law formalizes the concept of the directionality of thermodynamic processes, stating that heat cannot spontaneously flow from a colder body to a hotter body. This law introduces the idea of entropy, a measure of the disorder or randomness in a system, which tends to increase over time in isolated systems.

@@

@@

\begin{definition}[Equilibrium]
    An \textbf{ISOLATED} system is in equilibrium if its properties are uniform throughout the system and do not change with time (constant). \\
    (Example) A gas in a closed container that has been left undisturbed for a long time.
\end{definition}
\begin{example}
    Consider a gas in a cylinder with a movable piston. If the piston is held in place, the gas is in a steady state, as its properties (pressure, volume, temperature) do not change with time. However, if the piston is allowed to move, the gas will expand or compress, and its properties will change with time, indicating that it is not in a steady state. 
\end{example}
\begin{definition}[Process]
    A process is a transformation from one state to another. A process can be described by the path taken by the system in the property space. For example, a gas can undergo a process of compression or expansion, which changes its pressure, volume, and temperature.
\end{definition}

\subsection{Modeling Ideal Gas}
\begin{definition}[Quasi-Equilibrium Process (Reversible Path)]
    A quasi-equilibrium process is a process that occurs slowly enough that the system remains in a state of equilibrium at all times. In other words, the properties of the system change so (infinitely) slowly that they can be considered constant during the process. In this case, this path is reversible. \\
    (Example) A gas in a cylinder with a movable piston that is compressed or expanded very slowly, allowing the gas to remain in equilibrium throughout the process.
\end{definition}

\begin{definition}[Real Process (Irreversible Path)]
    A real process is a process that occurs at a finite rate, and the system may not remain in a state of equilibrium at all times. In other words, the properties of the system may change rapidly during the process, and the system may not be able to adjust to these changes quickly enough to maintain equilibrium. In this case, this path is irreversible. \\
    (Example) A gas in a cylinder with a movable piston that is compressed or expanded quickly, causing the gas to deviate from equilibrium during the process. In that process, the work done is more than that of a quasi-equilibrium process.
\end{definition}

\begin{shaded}
\paragraph{Motivation} We care about this because this is the process in which we ignite an engine, allowing gas to expand rapidly, pushing the piston down and doing work.
\end{shaded}

\begin{definition}[Ideal Gas Model]
    Consider an isolated system with molecules that:
    \begin{itemize}
        \item Have negligible volume compared to the volume of the container. (point mass)
        \item Do not interact with each other except during elastic collisions.
        \item Are in constant random motion without spin or vibration.
    \end{itemize}
    The best example of this is a noble gas (e.g., helium, neon, argon). Since they are inert and monoatomic. For all we care about in this course all gas are ideal unless:
    \begin{itemize}
        \item They are at very high pressure (Condier more then 1 atm).
        \item They are at very low temperature (Consider close to 0 K).
    \end{itemize}
\end{definition}

Such a system is called an ideal gas. The macroscopic observation of pressure is the average force per unit area exerted by the gas molecules colliding with the walls of the container, so if we increase the number of molecules or decrease the volume, the pressure increases proportionally, this is Boyle's law:

\begin{definition}[Boyle's Law]
    For a given amount of gas at constant temperature, the pressure of the gas is inversely proportional to its volume:
    \begin{equation}
        P \propto \frac{1}{V} \quad \Rightarrow \quad PV = \text{constant}
    \end{equation}
    where $N$ is the number of molecules. We can proof the $PV$ inverse relationship a momentum conservation argument. Consider linear path with lenght $L$, since collisions are elastic, the change in momentum for each collision is $\Delta p = 2mv_x$. The time between collisions is $\Delta t = \frac{2L}{v_x}$, so the force exerted on the wall is:
    $$
        F = \frac{\Delta p}{\Delta t} = \frac{2mv_x}{\frac{2L}{v_x}} = \frac{mv_x^2}{L}
    $$
    The pressure is then given by:
    $$
        P = \frac{F}{A} = \frac{mv_x^2}{AL} = \frac{mv_x^2}{V}
    $$
    Thus, we have $PV \propto mv_x^2$. Now, define the RMS velocity as $v_{rms} = \sqrt{\frac{1}{N} \sum_{i=1}^N v_i^2}$, and consider random motion, we have $\overline{v_x^2} = \overline{v_y^2} = \overline{v_z^2} = \frac{1}{3} v_{rms}^2$. Hence, we have:
    $$
        PV \propto Nm \overline{v_x^2} = \frac{1}{3} Nm v_{rms}^2
    $$
    where $Nm$ is the total mass of the gas. Then, we complete the arguement of the gas moving in all direction.
\end{definition}

\begin{example}[Boyle's Law]
    Consider a gas in a cylinder with a movable piston. If we compress the gas by pushing the piston down, the volume of the gas decreases, and the pressure increases. This relationship is described by Boyle's law, which states that for a given amount of gas at constant temperature, the pressure of the gas is inversely proportional to its volume:
\end{example}

We can also observe that increasing the temperature increases the average kinetic energy of the molecules, which increases the frequency and force of collisions with the walls of the container, leading to an increase in pressure, this is Charles's law:
\begin{definition}[Charles's Law, Total Internal Energy]
    For a given amount of gas at constant volume, the pressure of the gas is directly proportional to its absolute temperature:
    \begin{equation}
        P \propto T
    \end{equation}
    And the total internal energy of an ideal gas is the sum of the kinetic energies of all the gas molecules, which is also directly proportional to its absolute temperature:
    \begin{equation}
        U = \frac{3}{2} N k_B T = \frac{3}{2} n R_u T
    \end{equation}
    $U$ is a extensive property, so it is proportional to the number of molecules $N$ (or moles $n$). The intensive property is the specific internal energy:
    \begin{equation}
        u = \frac{U}{m} = \frac{3}{2} \frac{R_u}{M} T = \frac{3}{2} R T
    \end{equation}
    where $M$ is the molar mass of the gas, and
    \begin{equation}
        R = \frac{R_u}{M}
    \end{equation}
    is the specific gas constant.

    This is a consequence of the kinetic theory of gases, which model temperature as the average kinetic energy of the gas molecules, so we have:
    \begin{align*}
        \frac{1}{2} mv_{rms}^2 &= \frac{3}{2} R_u T \\
        \intertext{Dividing by $N_A$ (Avogadro's number), and define the boltzmann constant $k_B = \frac{R_u}{N_A}$, we have:}
        \frac{1}{2} m_e v_{rms}^2 &= \frac{3}{2} k_B T 
        \intertext{We multiply both sides by $N$ (number of molecules), and substitute into the previous equation, we have:}
        \frac{1}{2} m_e v_{rms}^2 N &= \frac{3}{2} k_B T N \\
    \end{align*}
    The above is the internal energy of the gas. Thus, we can deduce the following Ideal Gas Law, which matches macroscopic observations:
\end{definition}

\begin{definition}[Ideal Gas Law]
    Consider the combination of Boyle's and Charles's law, we have the ideal gas law:
    \begin{equation}
        PV = nR_uT = mRT
    \end{equation}
    where $R_u = 8.314\, \text{J/(mol K)}$ (or, equivalently, $R_u = 8 \times 10^3 \, \text{J/(kmol K)}$) is the universal gas constant, and $n$ is the number of moles of gas. $R = \frac{R_u}{M}$ is the specific gas constant, where $M$ is the molar mass of the gas.
\end{definition}

\begin{example}
    \textit{A spherical balloon with $d = 6\, m$ is filled with helium gas at $P = 200$ kPa and $T = 20^\circ C$. Calculate the mass and number of moles of helium in the balloon. (Assume ideal gas behavior).}

    We can use the ideal gas law to solve this problem. First, we need to calculate the volume of the balloon:
    $$        
    V = \frac{4}{3} \pi r^3 = \frac{4}{3} \pi (3)^3 = 113.1\, \text{m}^3
    $$
    Next, we can use the ideal gas law to calculate the number of moles of helium in the balloon:
    $$
        n = \frac{PV}{R_uT} = \frac{200 \times 10^3 \times 113.1}{8.314 \times (20 + 273)} = 9\,280\, \text{mol}
    $$
    Finally, we can calculate the mass of helium in the balloon using the molar mass of helium ($M = 4\, \text{g/mol}$):
    $$
        m = nM = 9\,280 \times 4  = 37\,120\, \text{g} = 37.12\, \text{kg}
    $$
\end{example}

\begin{example}[Adiabatic Compression]
    Consider an ideal gas undergoing adiabatic compression\footnote{This differs from \textbf{isothermal} compression, where the temperature remains constant.}. 
    Since no heat is exchanged with the surroundings ($Q=0$), the work done on the gas increases its internal energy, which raises the temperature. 
    Using the adiabatic relations,
    $$
        P V^{\gamma} = \text{constant}, 
        \quad T V^{\gamma - 1} = \text{constant},
    $$
    we see that a decrease in $V$ leads to an increase in both $P$ and $T$.
\end{example}

\begin{example}[Isothermal Expansion]
    Consider an ideal gas undergoing isothermal expansion\footnote{This differs from \textbf{adiabatic} expansion, where no heat is exchanged.}. 
    Since the temperature remains constant ($\Delta T = 0$), the internal energy of the gas does not change ($\Delta U = 0$). 
    The work done by the gas during expansion is equal to the heat absorbed from the surroundings ($Q = W$). 
    Using the ideal gas law, we can see that as $V$ increases, $P$ decreases, while $T$ remains constant.

\end{example}

\paragraph{Note} For all ideal gas processes, the internal energy $U$ depends only on temperature $T$, not on pressure $P$ or volume $V$. This is a key property of ideal gases and simplifies the analysis of thermodynamic processes involving them.

\begin{definition}[Change in Internal Energy]
    The change in internal energy of an ideal gas can be calculated using the specific heat capacity at constant volume ($c_v$):
    \begin{equation}
        \Delta U = \frac{3}{2} mk_B \Delta T = m c_v \Delta T
    \end{equation}
    where $c_v$ [J kg$^{-1}$ K$^{-1}$] is the specific heat capacity at constant volume. For a monoatomic ideal gas, we have:
    \begin{equation}
        c_v = \frac{3}{2} R
    \end{equation}
    where $R$ is the specific gas constant.
\end{definition}

\begin{example}
    \textit{How much energy is required to heat 10kg of air with 20$^\circ$C at 1 bar to 120$^\circ$ at constant volume? What is the final pressure? (Assume ideal gas behavior, and $c_{v, \text{air}} = 0.717$ kJ/kg K)}

    Consider the change of internal energy:
    $$
        \Delta U = m c_v \Delta T = 10 \times 100 \times c_{v, \text{air}} = 1\,000 \times 0.717 = 717\, \text{kJ}
    $$

    Now consider the change of pressure, we first isolate for the control:
    $$
        \frac{mR}{V} = \frac{P}{T}
    $$
    So we can compute the final pressure:
    $$
        \frac{P_1}{T_1} = \frac{P_2}{T_2} \Rightarrow P_2 = P_1 \frac{T_2}{T_1} = 1 \times \frac{393}{293} = 1.34\, \text{bar}
    $$
\end{example}

\section{First Law of Thermodynamics}
\begin{shaded}
\paragraph{Motivation} We want to understand how energy is conserved in a system, and how energy is transferred between a system and its surroundings. This is important for understanding how engines work, how to design efficient systems, and how to analyze thermodynamic processes.
\end{shaded}  
The same definition as before \ref{def:firstlaw}.

\begin{definition}[First Law of Thermodynamics for a Closed or Isolated System]
    For a closed system, the energy balance can be expressed as:
    \begin{equation}
        \Delta E_{in} - \Delta E_{out} = \Delta E_{system}
    \end{equation}
    where $\Delta E_{in}$ is the energy entering the system, $\Delta E_{out}$ is the energy leaving the system, and $\Delta E_{system}$ is the change in energy of the system. The total energy of the system can be expressed as:
    \begin{equation}
        E_{system} = U + KE + PE
    \end{equation}
    where $U$ is the internal energy, $KE$ is the kinetic energy, and $PE$ is the potential energy.
\end{definition}

\paragraph{Sign Convention} Relative to a system, we adopt the following sign convention:
\begin{itemize}
    \item Energy transfer to the system is positive.
    \item Energy transfer from the system is negative.
\end{itemize}

\begin{definition}[Energy Transfer from a System]
    The energy transfer from a system can occur in two ways: heat transfer ($Q$) and work transfer ($W$). The first law of thermodynamics can be expressed as:
    \begin{equation}
        dE = \delta Q + \delta W
    \end{equation}
\end{definition}

Dividing by $dt$, we have the rate form of the first law:
\begin{definition}[Rate Equation of the First Law of Thermodynamics]
    \begin{equation}
        \frac{dE}{dt} = \dot{Q} + \dot{W}
    \end{equation}
    where $\dot{Q}$ is the rate of heat transfer, and $\dot{W}$ is the rate of work transfer.
\end{definition}

\subsection{Work Modes}
\begin{definition}[Boundry Work]
    Force acts on the boundries of a systems and deformes them. \\
    (Example) Expansion and compression of a gas

    Assuming for a Quaso-equilibrium process, and no friction, we can compute the boundary work as:
    \begin{equation}
        \delta W = F dx  = P A dx = - P dV
    \end{equation}
    where $P$ is the pressure of the gas, $A$ is the area of the piston, and $dx$ is the displacement of the piston. The negative sign is because when the gas expands, it does work on the surroundings, so the work done by the system is negative. Also, we can consider the total work by:
    \begin{equation}
        w_{12} = -\int_{V_1}^{V_2} P dV
    \end{equation}
    \text{Note} Connecting that to the $PV$ diagram, the work done by the system is the area under the curve in the $PV$ diagram.
\end{definition}

\chapter{Heat Transfer}

### 2025-09-09 YEAR2/MAT292_LectureNotes/main.tex
sept 9

    To analyze the stability of the fixed points, we first find the fixed points by setting \( \frac{dx}{dt} = 0 \):
    $$\alpha x(1 - x) = 0 \implies x = 0 \text{ or } x = 1
    $$
    Next, we compute the derivative of \( f(x) = \alpha x(1 - x) \):
    $$\frac{df}{dx} = \alpha (1 - 2x)
    $$
    Evaluating this derivative at the fixed points, assumeing \( \alpha > 0 \):
    \begin{itemize}
        \item At \( x = 0 \):
        $$\frac{df}{dx} \bigg|_{x=0} = \alpha > 0 \quad \text{(unstable)}
        $$
        \item At \( x = 1 \):
        $$\frac{df}{dx} \bigg|_{x=1} = -\alpha < 0 \quad \text{(stable)}
        $$
    \end{itemize}
    Thus, the fixed point at \( x = 0 \) is unstable, while the fixed point at \( x = 1 \) is stable.
\end{example}

>>>>

    To analyze the stability of the fixed points, we first find the fixed points by setting \( \frac{dx}{dt} = 0 \):
    $$\alpha x(1 - x) = 0 \implies x = 0 \text{ or } x = 1
    $$
    Next, we compute the derivative of \( f(x) = \alpha x(1 - x) \):
    $$\frac{df}{dx} = \alpha (1 - 2x)
    $$
    Evaluating this derivative at the fixed points, assumeing \( \alpha > 0 \):
    \begin{itemize}
        \item At \( x = 0 \):
        $$\frac{df}{dx} \bigg|_{x=0} = \alpha > 0 \quad \text{(unstable)}
        $$
        \item At \( x = 1 \):
        $$\frac{df}{dx} \bigg|_{x=1} = -\alpha < 0 \quad \text{(stable)}
        $$
    \end{itemize}
    Thus, the fixed point at \( x = 0 \) is unstable, while the fixed point at \( x = 1 \) is stable.
\end{example}
\subsubsection{Integrating Factor Method for First-Order Linear ODEs}
\begin{definition}[First-Order Linear ODE]
    A first-order linear ordinary differential equation (ODE) is an equation of the form:
    \begin{equation}
        \frac{dy}{dx} + P(x)y = Q(x)
    \end{equation}
    where \( P(x) \) and \( Q(x) \) are functions of the independent variable \( x \).
\end{definition}

\begin{definition}[Integrating Factor Method]
    To solve a first-order linear ODE of the form:
    $$
        \frac{dy}{dx} + P(x)y = Q(x)
    $$
    we can use the integrating factor method. The steps are as follows:
    \begin{enumerate}
        \begin{subequations}

        \item Compute the integrating factor \( \mu(x) \):
        \begin{equation}
            \mu(x) = \exp\left( \int P(x) \, dx \right)
        \end{equation}
        \begin{shaded}
            This is from the fact that:
        \begin{align*}
            \frac{d\mu}{dx} &= P(x) \mu(x) \quad \text{(by requirement of step \ref{step:integrating-factor-multiplication})} \\
            \Rightarrow \frac{1}{\mu} \frac{d\mu}{dx} &= P(x) \\
            \Rightarrow \int \frac{1}{\mu} \frac{d\mu}{dx} \, dx &= \int P(x) \, dx \\
            \Rightarrow \ln|\mu| &= \int P(x) \, dx \\
            \Rightarrow \mu(x) &= \exp\left( \int P(x) \, dx \right)
        \end{align*}
        \end{shaded}
        \item Multiply both sides of the original ODE by the integrating factor: 
        \begin{equation}
            \mu(x) \frac{dy}{dx} + \mu(x) P(x)y = \mu(x) Q(x)
        \end{equation}
        \item Recognize that the left-hand side is the derivative of \( \mu(x)y \) \label{step:integrating-factor-multiplication}:
        \begin{equation}
            \frac{d}{dx}[\mu(x)y] = \mu(x) Q(x)
        \end{equation}
        \item Integrate both sides with respect to \( x \):
        \begin{equation}
            \mu(x)y = \int \mu(x) Q(x) \, dx + C
        \end{equation}
        where \( C \) is the constant of integration.
        \item Finally, solve for \( y(x) \):
        \begin{equation}
            y(x) = \frac{1}{\mu(x)}\left( \int \mu(x) Q(x) \, dx + C \right)
        \end{equation}
        \end{subequations}
    \end{enumerate}
\end{definition}

\begin{example}
    Consider the IVP involving a first-order linear ODE:
    $$
    y' + \frac{2}{t} y = \frac{\sin t}{t^2}, \quad y(\pi) = 1
    $$

    To solve this IVP using the integrating factor method, we first identify \( P(t) = \frac{2}{t} \) and \( Q(t) = \frac{\sin t}{t^2} \).

    Next, we compute the integrating factor:
    \begin{align*}
        \mu(t) &= \exp\left( \int \frac{2}{t} \, dt \right) \\
        &= \exp\left( 2 \ln |t| \right) \\
        &= |t|^2 = t^2
    \end{align*}

    Multiplying both sides of the original ODE by the integrating factor:
    $$
        t^2 y' + 2ty = \sin t
    $$

    Recognizing the left-hand side as the derivative of \( |t|^2 y \):
    \begin{align*}
        \frac{d}{dt}[t^2 y] &= \sin t \\
        t^2 y &= -\cos t + C \\
        y &= \frac{-\cos t + C}{t^2}
    \end{align*}

    To determine the constant \( C \), we use the initial condition \( y(\pi) = 1 \):
    \begin{align*}
        1 &= \frac{-\cos(\pi) + C}{\pi^2} \\
        \Rightarrow C &= \pi^2 - 1
    \end{align*}

    Thus, the solution to the IVP is:
    \begin{align*}
        y(t) &= \frac{-\cos t + \pi^2 - 1}{t^2}
    \end{align*}
\end{example}

\subsubsection{Exact ODEs}
\begin{definition}[Exact ODE]
    An ordinary differential equation (ODE) of the form:
    \begin{subequations}
    \begin{equation}
        M(x, y) + N(x, y) \frac{dy}{dx} = 0 \quad \text{or equivalently} \quad M(x, y) dx + N(x, y) dy = 0
    \end{equation}
    is said to be exact iff there exists a function \( \psi(x, y) \) such that:
    \begin{equation}
        M_y = \frac{\partial M}{\partial y} = N_x = \frac{\partial N}{\partial x}
    \end{equation}

    This implies a potential function \( \psi(x, y) \) exists such that:
    \begin{equation}
        M_y = \psi_{xy} = \psi_{yx} = N_x
    \end{equation}
    The general solution to the exact ODE is given by:
    \begin{equation}
        d\psi(x, y) = 0 \implies \psi(x, y) = C \quad (\text{By chain rule of $\psi$})
    \end{equation}
    \end{subequations}
    In this case, we can solve the ODE by finding the potential function \( \psi(x, y) \):
    \begin{equation}
        \psi(x, y) = \int M(x, y) \, dx + g(y)
    \end{equation}
    to find \( g(y) \), we can use the fact that:
    \begin{equation}
        \frac{\partial \psi}{\partial y} = \frac{\partial}{\partial y} \left( \int M(x, y) \, dx \right) = \int M_y \, dx = \int N_x \, dx = N(x, y)
    \end{equation}
    We can also solve this from the other direction. Or verify your answer this way.
\end{definition}

\begin{example}
    Consider the exact ODE:
    $$
    (3x^2y^2 + x + \cos{y}) \frac{dy}{dx} = -(2xy^3 + y)
    $$
    We identify \( N(x, y) = 3x^2y^2 + x + \cos{y} \) and \( M(x, y) = 2xy^3 + y \). So \( M_y = N_x = 6x^2y + 1 \), thus the ODE is exact. Then we compute the potential function:
    \begin{align*}
        \psi(x, y) &= \int M(x, y) \, dx \\
        &= \int (2xy^3 + y) \, dx \\
        &= x^2 y^3 + xy + g(y)
        \intertext{To find \( g(y) \), we compute:}
        \frac{\partial \psi}{\partial y} &= 3x^2 y^2 + x + g'(y) \\
        &= N(x, y) = 3x^2 y^2 + x + \cos{y} \\
        \Rightarrow g'(y) &= \cos{y} \\
        \Rightarrow g(y) &= \sin{y}
        \intertext{Thus, the potential function is:}
        \psi(x, y) &= x^2 y^3 + xy + \sin{y}
        \intertext{The general solution to the ODE is then:}
        \psi(x, y) &= C \\
        \Rightarrow x^2 y^3 + xy + \sin{y} &= C
    \end{align*}

    We can verify our answer by computing from the other direction:
    \begin{align*}
        \psi(x, y) &= \int N(x, y) \, dy \\
        &= \int (3x^2 y^2 + x + \cos{y}) \, dy \\
        &= x^2 y^3 + xy + \sin{y} + h(x)
        \intertext{To find \( h(x) \), we compute:}
        \frac{\partial \psi}{\partial x} &= 2x y^3 + y + h'(x) \\
        &= M(x, y) = 2xy^3 + y \\
        \Rightarrow h'(x) &= 0 \\
        \Rightarrow h(x) &= C'
        \intertext{Thus, we obtain the same potential function:}
        \psi(x, y) &= x^2 y^3 + xy + \sin{y} + C'
    \end{align*}
\end{example}
